Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2018-05-07 articles/formalism/sections/kernel_discussion.tex
updated figures, table of contents

We compute the Euler angles $(\alpha, \beta, \gamma)$ given the angular coordinates of any two Healpix pixels and use these to evaluate the convolution and the radiating kernels. To provide an intuition for how these kernels vary as a function of position of the central pixel we depict in \fig{fig:vis_kernel} the respective kernels at a few different locations on the sphere.
For illustration these functions are sampled at a very high Healpix resolution parameter of NSIDE=2048. All the plots have been rotated such that the central location marked by the black circle are in the centre of the figure. The horizontal and vertical lines that pass through the central black circle mark the local latitude and longitude respectively.

The real and imaginary part of the kernel $\mathcal{M}_G$ are identical irrespective of changes in the galactic latitude and longitude of the central pixel. This is consistent with the Green's function interpretation of these kernels. Note that these functions are not distorted when a part of the domain overlaps with the poles, as can be seen in the first three rows of \fig{fig:vis_kernel}. Both these facts can be associated with the fact that this function does not depend on the Euler angle $\gamma$. From \eq{eq:qu2eb_convolution_explicit} and \eq{eq:eb2qu_convolution_explicit} it is clear that $\mathcal{M}_r$ and $\mathcal{M}_i$ can be interpreted in the following ways,
%\revisit{Recall that the angle $\gamma$ defines the rotation about the final z-axis which co-alligns the planar axes. Since the function $\mathcal{M}$ does not depend of this final orientation, it implies a rotational invariance of the fields constructed by convolving with this kernel.}
%
\begin{subequations}
\beqry
\bmat E= -\mathcal{M}_r \\ B =+\mathcal{M}_i  \emat  \leftarrow \bmat Q=\delta(\hat{n} - \hat{n}_j)\\ U=0 \emat ~~;~~ \bmat E= -\mathcal{M}_i \\ B = - \mathcal{M}_r  \emat  \leftarrow \bmat Q=0 \\ U=\delta(\hat{n} - \hat{n}_j) \emat \,,  \\
\bmat Q= -\mathcal{M}_r \\ U = -\mathcal{M}_i  \emat  \leftarrow \bmat E=\delta(\hat{n} - \hat{n}_j)\\ B=0 \emat ~~;~~ \bmat Q= +\mathcal{M}_i \\ U = - \mathcal{M}_r  \emat  \leftarrow \bmat E=0 \\ B=\delta(\hat{n} - \hat{n}_j) \emat \,.
\eeqry
\end{subequations}
%
The kernels $\mathcal{D}$ \& $\mathcal{I}$ vary significantly as a function of galactic latitude of the central pixel as seen in the last four columns of \fig{fig:vis_kernel}. These kernels show a two fold symmetry in the vicinity of the poles and this arises due to Euler angle $\gamma \approx 0$ here and therefore $e^{i2(\alpha \pm \gamma)} \approx e^{i2\alpha}$. Note that in this region, the azimuthal profile of the real and imaginary part of these kernels  is similar to $\mathcal{M}_r$ and $\mathcal{M}_i$ respectively.  This also explains why the imaginary part of the band limited delta function $\mathcal{I}$ contributes just as much as the real part in these regions. On moving to lower latitudes, $\mathcal{D}$ quickly transitions to having a four fold symmetry while $\mathcal{I}$ transitions to being dominated by the real part and behaves more like the conventional delta function. This transition can be most easily understood in the flat sky limit where $\gamma = -\alpha$ which leads to the resultant 4 fold symmetry seen for $\mathcal{D}$ owing to $e^{i2(\alpha - \gamma)} =e^{i4\alpha}$ and $\mathcal{I}$ being dominated by the real part owing to $e^{-i2(\alpha + \gamma)} =1 + i0$. Since the flat sky approximation has most validity in the proximity of the equator these limiting tendencies of the respective kernels are seen in the bottom row of \fig{fig:vis_kernel} which depict the kernels evaluated at the equator $b=0^{\circ}$. The middle two row depict the kernels evaluated at a latitudes of $b=87^{\circ}~\&~ 80^{\circ}$ and serve to indicate the rate of this transition. These kernels are invariant under changes in longitude of the central pixel, the latitude being held fixed, as one may have expected.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quantifying the non-locality of E \& B modes} \label{sec:radial_locality}
%
\begin{figure}[t]
\centering
\includegraphics[width=0.8\columnwidth]{kernel/beta_kernel.pdf}
\caption{The figure depicts the radial part of the convolution kernels. These radial function have been evaluated with the band limit fixed at $\ell \in [2,24]$. The vertical dashed line marks the approximate Healpix pixel size of a NSIDE=8, which is the lowest resolution that allows access to $\ell_{\rm max}=24$.}
\label{fig:beta_kernel}
\end{figure}
%
In the previous section we presented a quantitative understanding of the  azimuthal dependence of the real space kernels, providing little insight into their radial dependence. The radial part of these kernels determines the non-locality of the respective operators and encodes all their multipole dependencies. For illustration the radial kernels ${_{\mm}f}, {_{\md}f} ~\&~ {_{\mi}f} $ are evaluated using the respective multipole sums given in \eq{eq:rad_ker_queb} and \eq{eq:f2_rad_ker} in the band limit $\ell \in [2,192]$ and the resultant functions are depicted in \fig{fig:beta_kernel}. 
%
\begin{figure}[t]
\centering
\subfigure{\includegraphics[width=0.96\columnwidth]{kernel/f_rad_ker_fn_of_ellmax.pdf}}
\subfigure{\includegraphics[width=0.48\columnwidth]{kernel/fm2_rad_ker_fn_of_ellmax.pdf}}
\subfigure{\includegraphics[width=0.48\columnwidth]{kernel/fp2_rad_ker_fn_of_ellmax.pdf}}
\caption{The top panel depicts the radial function  ${_{\mm}f}(\beta,\ell_{\rm min},\ell_{\rm max})$ while the bottom left and right panels show the radial functions ${}_{\mi}f(\beta,\ell_{\rm min},\ell_{\rm max}) ~\&~ {}_{\md}f(\beta,\ell_{\rm min},\ell_{\rm max})$ respectively, for fixed $\ell_{\rm min}=2$ and varying $\ell_{\rm max}$ as indicated by their legends. All the curves are normalized such that their maxima is set to unity. The horizontal solid black line marks the location where the amplitude of the respective kernels fall below 1\% of its maximum. The slanted dashed gray lines indicate a power law fit (by eye) to the envelope of the radial functions. The black short vertical dashed lines indicate the transition points as predicted by the empirically derived relation for the non-locality parameter $\beta_0=min(180,180 \frac{24}{\ell_{\rm max}})$.}
%While the envelopes for function ${_{\mm}f}(\beta)~\&~ {}_{\mi}f(\beta)$ are fit well by the power law $\propto \beta^{-1.5}$, the envelope for the function ${}_{\md}f(\beta)$ is seen to have a slightly steeper slope $\propto \beta^{-1.65}$.}
\label{fig:rad_ker_decay}
\end{figure}
%

@@

It is useful to define a characteristic radius of the region from which the resultant fields evaluated at any location get most of their contribution.  Our primary interest is in quantifying the non-locality of the scalar modes $E$ \& $B$, therefore we define the abscissa at which the function ${_{\mm}f}(\beta,\ell_{\rm min}=2,\ell_{\rm max})$ transits to being monotonously below 1\% of the maxima of the function as the non-locality parameter: $\beta_{o}$.
%For $\ell_{\rm max}=24$, the maximum multipole accessible on a Nside=8 Healpix map, the non-locality parameter $\beta_0=180^{\circ}$ as the radial function never falls monotonously below 1\% of its global maxima as seen in \fig{}. Using this fact and the self similar property of the radial functions, we define the following empirical relation: $\beta_o= {\rm min}(180,180 \frac{24}{\ell_{\rm max}})$, as a means of estimating the non-locality parameter given the maximum multipole $\ell_{\rm max}$ accessible for analysis.
The simple empirical derived relation: $\beta_o= {\rm min}(180,180 \frac{24}{\ell_{\rm max}})$ provides a reasonable estimate of this transition point for  ${}_{\mm}f$. This empirical function also provides a reasonable estimate for the transition point for the function ${}_{\md}f$ particularly for high $\ell_{\rm max}$ and it systematically over-predicts the transition point for the function ${}_{\mi}f$ as seen in \fig{fig:rad_ker_decay}.

>>>>

We compute the Euler angles $(\alpha, \beta, \gamma)$ given the angular coordinates of any two Healpix pixels and use these to evaluate the convolution and the radiating kernels. To provide an intuition for how these kernels vary as a function of position of the central pixel we depict in \fig{fig:vis_kernel} the respective kernels at a few different locations on the sphere.
While the kernels are evaluated in the band limit $\ell \in [2,192]$, for illustration these functions are sampled at a very high Healpix resolution parameter of NSIDE=2048. All the plots have been rotated such that the central location marked by the black circle are in the centre of the figure. The horizontal and vertical lines that pass through the central black circle mark the local latitude and longitude respectively.

The real and imaginary part of the kernel $\mathcal{M}_G$ are identical irrespective of changes in the galactic latitude and longitude of the central pixel. This is consistent with the Green's function interpretation of these kernels. Note that these functions are not distorted when a part of the domain overlaps with the poles, as can be seen in the first three rows of \fig{fig:vis_kernel}. Both these facts can be associated with the fact that this function does not depend on the Euler angle $\gamma$. From \eq{eq:qu2eb_convolution_explicit} and \eq{eq:eb2qu_convolution_explicit} it is clear that $\mathcal{M}_r$ and $\mathcal{M}_i$ can be interpreted in the following ways,
%\revisit{Recall that the angle $\gamma$ defines the rotation about the final z-axis which co-alligns the planar axes. Since the function $\mathcal{M}$ does not depend of this final orientation, it implies a rotational invariance of the fields constructed by convolving with this kernel.}
%
\begin{subequations}
\beqry
\bmat E= -\mathcal{M}_r \\ B =+\mathcal{M}_i  \emat  \leftarrow \bmat Q=\delta(\hat{n} - \hat{n}_j)\\ U=0 \emat ~~;~~ \bmat E= -\mathcal{M}_i \\ B = - \mathcal{M}_r  \emat  \leftarrow \bmat Q=0 \\ U=\delta(\hat{n} - \hat{n}_j) \emat \,,  \\
\bmat Q= -\mathcal{M}_r \\ U = -\mathcal{M}_i  \emat  \leftarrow \bmat E=\delta(\hat{n} - \hat{n}_j)\\ B=0 \emat ~~;~~ \bmat Q= +\mathcal{M}_i \\ U = - \mathcal{M}_r  \emat  \leftarrow \bmat E=0 \\ B=\delta(\hat{n} - \hat{n}_j) \emat \,.
\eeqry
\end{subequations}
%
The kernels $\mathcal{D}$ \& $\mathcal{I}$ vary significantly as a function of galactic latitude of the central pixel as seen in the last four columns of \fig{fig:vis_kernel}. These kernels show a two fold symmetry in the vicinity of the poles and this arises due to Euler angle $\gamma \approx 0$ here and therefore $e^{i2(\alpha \pm \gamma)} \approx e^{i2\alpha}$. Note that in this region, the azimuthal profile of the real and imaginary part of these kernels  is similar to $\mathcal{M}_r$ and $\mathcal{M}_i$ respectively.  This explains why the imaginary part of the band limited delta function $\mathcal{I}$ contributes just as much as the real part in these regions. On transiting to lower latitudes, $\mathcal{D}$ quickly transitions to having a four fold symmetry while $\mathcal{I}$ transitions to being dominated by the real part and behaves more like the conventional delta function. This transition can be most easily understood in the flat sky limit where $\gamma = -\alpha$ which leads to the resultant 4 fold symmetry seen for $\mathcal{D}$ owing to $e^{i2(\alpha - \gamma)} =e^{i4\alpha}$ and $\mathcal{I}$ being dominated by the real part owing to $e^{-i2(\alpha + \gamma)} =1 + i0$. Since the flat sky approximation has most validity in the proximity of the equator these limiting tendencies of the respective kernels are seen in the bottom row of \fig{fig:vis_kernel} which depict the kernels evaluated at the equator $b=0^{\circ}$. The middle two row depict the kernels evaluated at a latitudes of $b=87^{\circ}~\&~ 80^{\circ}$ and serve to indicate the rate of this transition. These kernels are invariant under changes in longitude of the central pixel, the latitude being held fixed, as one may have expected.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Quantifying the non-locality of E \& B modes} \label{sec:radial_locality}
\subsection{The non-locality of the real space operators} \label{sec:radial_locality}
%
\begin{figure}[t]
\centering
\includegraphics[width=0.8\columnwidth]{beta_kernel.pdf}
\caption{The figure depicts the radial part of the convolution kernels. These radial function have been evaluated with the band limit fixed at $\ell \in [2,24]$. The vertical dashed line marks the approximate Healpix pixel size of a NSIDE=8, which is the lowest resolution that allows access to $\ell_{\rm max}=24$.}
\label{fig:beta_kernel}
\end{figure}
%
In the previous section we presented a quantitative understanding of the  azimuthal dependence of the real space kernels, providing little insight into their radial dependence. The radial part of these kernels determines the non-locality of the respective operators and encodes all their multipole dependencies. For illustration the radial kernels ${_{\mm}f}, {_{\md}f} ~\&~ {_{\mi}f} $ are evaluated using the respective multipole sums given in \eq{eq:rad_ker_queb} and \eq{eq:f2_rad_ker} in the band limit $\ell \in [2,192]$ and the resultant functions are depicted in \fig{fig:beta_kernel}. 
%
\begin{figure}[t]
\centering
\subfigure{\includegraphics[width=0.96\columnwidth]{rad_ker_fn_of_ellmax.pdf}}
\subfigure{\includegraphics[width=0.48\columnwidth]{rad_ker_i_fn_of_ellmax.pdf}}
\subfigure{\includegraphics[width=0.48\columnwidth]{rad_ker_d_fn_of_ellmax.pdf}}
\caption{The top panel depicts the radial function  ${_{\mm}f}(\beta,\ell_{\rm min},\ell_{\rm max})$ while the bottom left and right panels show the radial functions ${}_{\mi}f(\beta,\ell_{\rm min},\ell_{\rm max}) ~\&~ {}_{\md}f(\beta,\ell_{\rm min},\ell_{\rm max})$ respectively, for fixed $\ell_{\rm min}=2$ and varying $\ell_{\rm max}$ as indicated by their legends. All the curves are normalized such that their maxima is set to unity. The horizontal solid black line marks the location where the amplitude of the respective kernels fall below 1\% of its maximum. The thin slanted dashed gray lines indicate a power law fit (by eye) to the envelope of the radial functions. The thick black short vertical dashed lines indicate the transition points as predicted by the empirically derived relation for the non-locality parameter $\beta_0=min(180,180 \frac{22}{\ell_{\rm max}})$.}
%While the envelopes for function ${_{\mm}f}(\beta)~\&~ {}_{\mi}f(\beta)$ are fit well by the power law $\propto \beta^{-1.5}$, the envelope for the function ${}_{\md}f(\beta)$ is seen to have a slightly steeper slope $\propto \beta^{-1.65}$.}
\label{fig:rad_ker_decay}
\end{figure}
%

@@

It is useful to define a characteristic radius of the region from which the evaluated resultant fields get most of their contribution.  Our primary interest is in quantifying the non-locality of the scalar modes $E$ \& $B$, therefore we define the abscissa at which the function ${_{\mm}f}(\beta,\ell_{\rm min}=2,\ell_{\rm max})$ transits to being monotonously below 1\% of the maxima of the function as the non-locality parameter: $\beta_{o}$.
%For $\ell_{\rm max}=24$, the maximum multipole accessible on a Nside=8 Healpix map, the non-locality parameter $\beta_0=180^{\circ}$ as the radial function never falls monotonously below 1\% of its global maxima as seen in \fig{}. Using this fact and the self similar property of the radial functions, we define the following empirical relation: $\beta_o= {\rm min}(180,180 \frac{24}{\ell_{\rm max}})$, as a means of estimating the non-locality parameter given the maximum multipole $\ell_{\rm max}$ accessible for analysis.
The simple empirical derived relation: $\beta_o= {\rm min}(180,180 \frac{22}{\ell_{\rm max}})$ provides a reasonable estimate of this transition point for  ${}_{\mm}f$. This empirical function also provides a reasonable estimate for the transition point for the function ${}_{\md}f$ particularly for high $\ell_{\rm max}$ and it systematically over-predicts the transition point for the function ${}_{\mi}f$ as seen in \fig{fig:rad_ker_decay}.

### 2018-05-07 articles/formalism/sections/real_space_formalism.tex
updated figures, table of contents

%\section{Real space polarization operators}
\section{Polarization primer}
%\textit{Alternative for bringing the heuristic argument earlier.}

@@

%--------------------------------------------------------
\subsection{Standard $E$/$B$ fields and Matrix notation} \label{sec:mat_pol_intro}

@@

Our derivation of real space operators is more transparent in a matrix-vector notation.\footnote{While we work with the matrix and vector sizes given in terms of some pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}
We introduce a matrix that encodes spin spherical harmonic basis vectors,
%
\beq
{}_{|s|}\mathcal{Y}= \bmat _{+s}Y & 0 \\ 0 & _{-s}Y \emat _{2 \rm N_{\rm pix} \times 2 \rm N_{\rm alms}} \,%,
\eeq
%
%where $s$ denotes the spin of the basis functions.
We will be working with cases $s \in [0,2]$. Each column maps to a specific harmonic basis function (i.e. indexed by $\ell m$) and each row maps to a pixel on the sphere. This matrix is not square in general: the number of rows is determined by the pixelization and the number of columns is set by the number of basis functions (e.g. determined by the band limit).

>>>>

\section{Polarization primer}
\subsection{A heuristic argument for constructing $E/B$ fields on the sphere}
\comment{Need a better title for the section that gives the heuristic argument}
%\textit{Alternative for bringing the heuristic argument earlier.}

@@

%--------------------------------------------------------
\subsection{Standard $E$/$B$ fields} \label{sec:mat_pol_intro}

@@

\section{Real space polarization operators}
\subsection{Matrix notation} \label{sec:mat_pol_intro}
Our derivation of real space operators is more transparent in a matrix-vector notation.\footnote{While we work with the matrix and vector sizes given in terms of some pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}
We introduce a matrix that encodes spin spherical harmonic basis vectors,
%
\beq
{}_{|s|}\mathcal{Y}= \bmat _{+s}Y & 0 \\ 0 & _{-s}Y \emat _{2 \rm N_{\rm pix} \times 2 \rm N_{\rm alms}} \,%,
\eeq
%
%where $s$ denotes the spin of the basis functions.
We will be working with cases $s \in [0,2]$. Each column maps to a specific harmonic basis function (i.e. indexed by $\ell m$) and each row maps to a pixel on the sphere. This matrix is not square in general: the number of rows is determined by the pixelization and the number of columns is set by the number of basis functions (e.g. determined by the band limit).
